---
title: "Simulating glaciation with ms"
author: Peter Ralph
date: May 5, 2015
---


```{r setup, include=FALSE}
require(ape)
source("msarg.R")
fig.dim <- 5
knitr::opts_chunk$set(fig.height=fig.dim,fig.width=2*fig.dim,fig.align='center')
```

**Scenario:** 
A single lattice is hit by a "glacier" and contracts into 2 isolated subpopulations, 
which after some time of isolation, subsequently expand out into the rest of the grid, and come into contact.

A demographic model is a list of population models on the same grid, with a bit more information (time to switch for each).
The way to build a demographic model is to:

1. Start with a grid,
2. then moving backwards in time, copy the last state, 
3. then modify it.

The function `add_to_demography` does this copy-and-modify step;
and demographic models can be modified directly by subsetting.


Building the demographic model
==============================

Initialization
--------------

First, let's choose the parameters for the grid:
```{r params}
nrow <- 100               # width of grid
ncol <- 40                # height of grid
glacier.width <- 70        # with of the glacier (in the middle)
mig.rate <- 1             # migration rate between neighbors in the grid
growth.time <- 200          # units of time (in Ne) it took to re-expand post-glacier
glacier.end <- 500          # units of time ago (in Ne) that glaciation ended
glacier.begin <- 1000        # units of time ago (in Ne) that glaciation began
```
Note that `Ne` is actually the number of individuals *per deme*, so the actual total effective population size is much larger, `r nrow*ncol`.

Now, set up the grid,
and use it to initialize a `demography` object.
```{r init_grid}
dem <- demography( grid_array(nlayers=1,nrow=nrow,ncol=ncol,N=1,mig.rate=mig.rate) )
plot(dem[[1]])
```

Create the glacier
------------------

We'll first create the "glaciated" state,
then use a helper function to interpolate between that and the full state.
```{r glaciation}
mask <- matrix(1,nrow=nrow(dem),ncol=ncol(dem))
mask[ abs(row(mask)-nrow(dem)/2) < glacier.width/2] <- 0
dem <- add_to_demography( dem, tnew=glacier.end, fn=modify_grid_layer, layer=1, dN=mask )
plot(dem[[2]])
```

Postglacial expansion
---------------------

The function `logistic_interpolation` interpolates between two population states
using local density-regulated population growth dynamics:
if the population size in deme $(i,j)$, $t$ units of time ago, was $N_{i,j}(t)$,
then one time step closer to the present it is
$$ 
N_{i,j}(t-dt) = N_{i,j}(t) 
    + dt \times r N_{i,j}(t) ( N_{i,j}^* - N_{i,j}(t) ) 
    + dt \times \frac{\sigma}{8} ( N_{i+1,j}(t) + N_{i-1,j}(t) + N_{i,j+1}(t) + N_{i,j-1}(t) - 4 N_{i,j}(t) )
.  
$$
Here $\sigma$ is the migration rate, $r$ is the replacement rate, and $N_{i,j}^*$ is the carrying capacity;
this generates a traveling wave that moves at speed $c = \sigma \sqrt{2r}$,
and has width of order $\sigma / \sqrt{r}$.

You specify the number of time steps you want `ms` to see of this process;
the function `logistic_interpolation` does a finer-grained version of this to get a nice smooth spread.
We could tell it $\sigma$ and $r$, but it is easier to get it to match a desired scenario
by specifying the speed and width. 
Note that at the end of the interpolation, the population goes instantly to its final state,
so if you don't get the speed right it will either expand too quick or jump suddenly at the end.
Also note that the wave takes some time to get going, so to make it arrive 
you will likely have to make it move faster than a naive calculation would suggest
to get it to arrive in time.
```{r expansion}
speed <- 1.2 * ( glacier.width ) / growth.time 
width <- 10
dem <- logistic_interpolation( dem, 
    t.end=glacier.end-growth.time,   # time ago expansion ended
    t.begin=glacier.end,             # time ago exapnsion began
    nsteps=10,                        # number of time steps to pass to ms
    speed=speed, width=width )       # parameters; could also pass sigma= and r= .
```

The arrival of the glacier
--------------------------

Now we just need to add the pre-glacier scenario,
which is the same as the final state:
```{r initial_state}
dem <- add_to_demography( dem, tnew=glacier.begin, pop=1 )
```

What it looks like
------------------

Here's what the sequence looks like:
(omitting arrows for clarity)
```{r plot_model}
plot(dem,do.arrows=FALSE)
```


Simulating with ms
==================

Trees, in R
-----------

Since we've only got one, unchanging, demographic scenario,
we don't have to make a special `demography` object.
We'll use a convenience function to pick some random sampling locations,
two each from 30 random locations.
```{r sample_config}
sample.config <- sample_locations( dem, n=30, each=2 )
sample.cols <- rainbow(nrow(sample.config))
sample.config
```
With this in hand, we can run `ms`.
We'll ask it to output trees, not sequence.
For some reason, it gives an error if we keep it all in `R`: 
maybe the string of parameters is too long?
(Using `tofile=TRUE` puts the parameters in a file, passed to `ms` via `-f`.)
```{r tree_msarg}
ms.output <- run_ms( dem, nsamp=sample.config, trees=TRUE, nreps=3 )
tree.output <- trees_from_ms( file.path(ms.output,"msoutput.txt") )
layout(t(1:2))
for (tree in tree.output) {
    plot_sample_config( dem, sample.config )
    abline(v=(nrow(dem)+c(1,-1)*glacier.width)/2, lty=2, lwd=2, col='grey')
    ape::plot.phylo( tree, tip.color=sample.cols[findInterval(as.numeric(tree$tip.label),1+c(0,cumsum(sample.config[,4])))] )
    ape::axisPhylo(1)
    abline(v=get("last_plot.phylo",envir=.PlotPhyloEnv)$x.lim[2]-c(glacier.end,glacier.begin), lty=2, lwd=2, col='grey')  # grrr, ape::
}
```

Sequence, to a file
-------------------

To get sequence, we could do as follows,
this time outputting to a directory:
```{r seq_msarg}
ms.output <- run_ms( dem, nsamp=sample.config, theta=0.001 )
cat(paste(scan(file.path(ms.output,"msoutput.txt"),what='char',sep='\n'),collapse='\n'))
```
