---
title: "Using msarg to construct ms models of migration."
author: Peter Ralph
date: May 1, 2015
---

```{r setup, include=FALSE}
source("msarg.R")
fig.dim <- 5
knitr::opts_chunk$set(fig.height=fig.dim,fig.width=2*fig.dim,fig.align='center')
```

The basic object is the population model for a given chunk of time.
We'll just deal with sets of populations on a grid,
so that will be a `gridArray` object.
Let's make one, 
with two layers,
of a 30 x 20 grid,
within-layer migration rates of 1 and between-layer rates of 0.1:
```{r init_grid, fig.height=2*fig.dim}
ga <- grid_array(nlayers=2,nrow=30,ncol=20,N=1000,mig.rate=1,admix.rate=0.1)
plot(ga)
```
In the plot, the diagonal shows arrows weighted by migration rates within a given layer,
and offdiagonals show circles with size proportional to the "admixture" migration between the relevant layers.
This looks pretty boring because all the migration rates are the same,
and isn't showing between-layer migration rates at the moment.
So we have something to look at, let's 
increase migration rates on one layer,
and then restrict each layer to a circle
centered on the two corners:
```{r zero_circles, fig.height=2*fig.dim}
ga <- modify_migration(ga,layer.ind=1,doutM=rexp(ncol(ga)),dinM=rexp(nrow(ga)))
plot(ga)
ga <- restrict_patch(ga,layer.ind=1,xy=c(1,1),r=10)
ga <- restrict_patch(ga,layer.ind=2,xy=c(nrow(ga),ncol(ga)),r=10)
plot(ga)
```
Note that each population also has an associated "exponential growth" rate.

Now, here's an example of stringing these together into a model,
with two time steps,
and a `dt` of 1 between them.
```{r init_dem}
dem <- pop_to_dem( grid_array(nlayers=2,nrow=30,ncol=20,N=1000,mig.rate=1,admix.rate=0.1) )
dem[[1]] <- modify_migration(dem[[1]],layer.ind=1,doutM=rexp(ncol(dem[[1]])),dinM=rexp(nrow(dem[[1]])))
dem <- add_to_demography( dem, dt=1, restrict_patch, layer.ind=1,xy=c(1,1),r=10 )
dem[[2]] <- restrict_patch(dem[[2]],layer.ind=2,xy=c(nrow(dem[[1]]),ncol(dem[[1]])),r=10)
cat("t=",dem@t[1],"\n")
plot(dem[[1]])
cat("t=",dem@t[2],"\n")
plot(dem[[2]])
```

We also have a function to convert these to ms command line parameters.
To check this, let's try a smaller model.
```{r ms_args}
dem <- pop_to_dem( grid_array(nlayers=2,nrow=3,ncol=2,N=1000,mig.rate=1,admix.rate=0.1) )
dem[[1]] <- modify_migration(dem[[1]],layer.ind=1,doutM=rexp(ncol(dem[[1]])),dinM=rexp(nrow(dem[[1]])))
dem <- add_to_demography( dem, dt=1, restrict_patch, layer.ind=1,xy=c(1,1),r=1 )
dem[[2]] <- restrict_patch(dem[[2]],layer.ind=2,xy=c(nrow(dem[[1]]),ncol(dem[[1]])),r=1)
print( msarg(dem,theta=1000) )
```

