---
title: "Simulating a barrier with ms"
author: Peter Ralph
date: May 4, 2015
---


```{r setup, include=FALSE}
require(ape)
library(msarg)
fig.dim <- 5
knitr::opts_chunk$set(fig.height=fig.dim,fig.width=2*fig.dim,fig.align='center')
```

**Scenario:** 
Two halves of a lattice that are divided at some point in the past by a barrier to migration, 
and then, closer to the present, that barrier is removed and they start to flow back into one another.

A demographic model is a list of population models on the same grid, with a bit more information (time to switch for each).
The way to build a demographic model is to:

1. Start with a grid,
2. then moving backwards in time, copy the last state, 
3. then modify it.

The function `add_to_demography` does this copy-and-modify step;
and demographic models can be modified directly by subsetting.


Building the demographic model
==============================

Initialization
--------------

First, let's choose the parameters for the grid:
```{r params}
nrow <- 30                # width of grid
ncol <- 20                # height of grid
barrier.time <- 3         # time ago the barrier ended
rejoin.time <- 5          # time ago the barrier began
barrier.rows <- c(10,20)  # locations of the barrier
```

Now, we will initialize the demographic model with a grid:
```{r init_grid}
dem <- demography( grid_array(nlayers=1,nrow=nrow,ncol=ncol,N=1,mig.rate=1) )
plot(dem[[1]])
```

Add the barrier
---------------

Then, we copy-and-modify that first state, adding it to the demography:
`add_to_demography` applies the function specified by `fn=` to a population state
(by default the most recent one), with the extra parameters given,
and appends it to the demography `dem`, at time given by `tnew`:
```{r add_barrier}
dem <- add_to_demography( dem, tnew=barrier.time, fn=add_barrier, layer=1, rows=barrier.rows )
plot(dem[[2]])
```

Remove the barrier
------------------

To remove the barrier, we'll just set the state back to what it was before (this is `pop[[1]]`, the first state):
```{r remove_barrier}
dem <- add_to_demography( dem, tnew=rejoin.time, pop=dem[[1]] )
plot(dem[[3]])
```


Simulating with ms
==================

Trees, in R
-----------

To turn this into the command-line parameters for `ms`, we first turn it into an `msarg` object,
which is just a named list,
but nicer to parse with `R` than just a big long string.
This also needs to know now many samples we will take, and where (as `ms` does).
`ms` needs the sample configuration as a giant long vector, 
but we keep track of it in a four-column matrix, (row, column, layer, number of samples).
```{r sample_config}
sample.config <- cbind(
    row=c(5,5,25,25),
    col=c(5,15,5,15),
    layer=c(1,1,1,1),
    n=c(2,2,2,2)
    )
sample.cols <- rainbow(nrow(sample.config))
sample.config
```
With this in hand, we can run `ms`.
We'll ask it to output trees, not sequence,
and keep it all internal to `R` (by default it outputs to a new directory):
```{r tree_msarg}
ms.output <- suppressWarnings( run_ms( dem, nsamp=sample.config, trees=TRUE, tofile=FALSE, nreps=3 ) ) # never mind these warnings
tree.output <- trees_from_ms( ms.output )
layout(t(1:2))
for (tree in tree.output) {
    plot_sample_config( dem, sample.config )
    abline(v=barrier.rows, lty=2, lwd=2, col='grey')
    ape::plot.phylo( tree, tip.color=sample.cols[findInterval(as.numeric(tree$tip.label),1+c(0,cumsum(sample.config[,4])))] )
    ape::axisPhylo(1)
}
```

Sequence, to a file
-------------------

To get sequence, we could do as follows,
this time outputting to a directory:
```{r seq_msarg}
ms.output <- run_ms( dem, nsamp=sample.config, theta=0.01 )
cat(paste(scan(file.path(ms.output,"msoutput.txt"),what='char',sep='\n'),collapse='\n'))
```
