---
title: "Simulating parallel universes with ms"
author: Peter Ralph
date: May 4, 2015
---


```{r setup, include=FALSE}
require(ape)
library(msarg)
fig.dim <- 5
knitr::opts_chunk$set(fig.height=fig.dim,fig.width=2*fig.dim,fig.align='center')
```

**Scenario:** 
Parallel "universe" lattices with some migration between "universes" in a zone of "universe" overlap.

A demographic model is a list of population models on the same grid, with a bit more information (time to switch for each).
The way to build a demographic model is to:

1. Start with a grid,
2. then moving backwards in time, copy the last state, 
3. then modify it.

The function `add_to_demography` does this copy-and-modify step;
and demographic models can be modified directly by subsetting.


Building the demographic model
==============================

Initialization
--------------

First, let's choose the parameters for the grid:
```{r params}
nrow <- 100               # width of grid
ncol <- 40                # height of grid
overlap.size <- 20        # with of the overlap (in the middle)
admix.rate <- 0.05
mig.rate <- 1
```

Now, set up the grid.
```{r init_grid}
ga <- grid_array(nlayers=2,nrow=nrow,ncol=ncol,N=1,mig.rate=mig.rate,admix.rate=admix.rate) 
plot(ga)
```

Create the overlap
------------------

Then, we modify that first state, by zeroing out populations in each layer where we don't want them.
We do this by creating 'masks' for each and using `modify_grid_layer`, which multiplies `N` by the mask we pass in with `dN`:
```{r modify_layers}
mask.1 <- mask.2 <- matrix(1,nrow=nrow(ga),ncol=ncol(ga))
mask.1[row(mask.1) > nrow(ga)/2+overlap.size/2] <- 0
mask.2[row(mask.1) < nrow(ga)/2-overlap.size/2] <- 0
ga <- modify_grid_layer( ga, layer=1, dN=mask.1 )
ga <- modify_grid_layer( ga, layer=2, dN=mask.2 )
plot(ga)
```

Simulating with ms
==================

Trees, in R
-----------

Since we've only got one, unchanging, demographic scenario,
we don't have to make a special `demography` object.
We'll use a convenience function to pick some random sampling locations,
two each from 30 random locations.
```{r sample_config}
sample.config <- sample_locations( ga, n=30, each=2 )
sample.cols <- rainbow(nrow(sample.config))
sample.config
```
With this in hand, we can run `ms`.
We'll ask it to output trees, not sequence.
For some reason, it gives an error if we keep it all in `R`: 
maybe the string of parameters is too long?
(Using `tofile=TRUE` puts the parameters in a file, passed to `ms` via `-f`.)
```{r tree_msarg}
ms.output <- run_ms( ga, nsamp=sample.config, trees=TRUE, nreps=3 )
tree.output <- trees_from_ms( file.path(ms.output,"msoutput.txt") )
layout(t(1:2))
for (tree in tree.output) {
    plot_sample_config( ga, sample.config )
    abline(v=(nrow(ga)+c(1,-1)*overlap.size)/2, lty=2, lwd=2, col='grey')
    ape::plot.phylo( tree, tip.color=sample.cols[findInterval(as.numeric(tree$tip.label),1+c(0,cumsum(sample.config[,4])))] )
    ape::axisPhylo(1)
}
```

Sequence, to a file
-------------------

To get sequence, we could do as follows,
this time outputting to a directory:
```{r seq_msarg}
ms.output <- run_ms( ga, nsamp=sample.config, theta=0.01 )
cat(paste(scan(file.path(ms.output,"msoutput.txt"),what='char',sep='\n'),collapse='\n'))
```
