---
title: "Using msarg to construct ms models of migration."
author: Peter Ralph
date: May 1, 2015
---

```{r setup, include=FALSE}
require(ape)
library(msarg)
fig.dim <- 5
knitr::opts_chunk$set(fig.height=2*fig.dim,fig.width=2*fig.dim,fig.align='center')
```

Here are some simpler, probably better, specific examples:

1. Appearance, and disappearance, of [a barrier](barrier-howto.html)
2. [Parallel universes](parallel-universes-howto.html)
3. Sudden appearance, and gradual retreat, of [a glacier](glaciation-howto.html)


Forwards and Reverse Time
=========================

The migration rates of `ms` are:
"$m_{ij}$ the fraction of subpopulation $i$ which is made up of migrants from subpopulation $j$ each generation".
But to translate to more natural units of migration,
we need to scale these rates by population size;
as the `ms` documentation notes:
adding `-n 3 0.5` has the effect that:
"the number of migrant copies arriving each generation
from a particular population, or from all other subpopulations, to population 3
is half as large as a consequence of this command. This is because the fraction
of migrants is the same, but the total number of individuals is half what it
would have been without the -n command."
In other words, the `-m` of `ms` gives the probability of migration of a *lineage*, in reverse time,
not an individual.

To make this more intuitive,
and along the way make it easier to specify range expansions,
we're going to be specifying migration rates in *forwards time*.
The migration rate $m_{ij}$ of `ms` is:
$$ m_{ij} = \frac{ \text{(number of inmigrants to $i$ from $j$ per generation)} }{ \text{(size of population $i$)} } $$
or equivalently
$$ m_{ij} = \text{(probability an indiv from $j$ will migrate to $i$ per generation)} \times \frac{ \text{(size of population $j$)} }{ \text{(size of population $i$)} } $$
We'll specify things in terms of *probability an indiv from $j$ will migrate to $i$ per generation*,
which we call $\tilde m_{ji}$,
and then when outputting to `ms` format, 
compute $m_{ij} =  \tilde m_{ji} n_j/n_i$.

After this, as populations get smaller (back in time),
the lineages tend to move out of them;
so rather than dissappearing populations (with `-ej` and the like)
we can just make them very small, and lineages will instantly do the right thing.


Population growth
=================

`ms` allows specification of population growth rates.
These don't interact with migration in the way we'd want them to (see above);
so we won't use this option.


Grid Arrays
===========

The basic object is the population model for a given chunk of time.
We'll just deal with sets of populations on a grid,
so that will be a `gridArray` object.
Let's make one, 
with two layers,
of a 30 x 20 grid,
within-layer migration rates of 1 and between-layer rates of 0.1:
```{r init_grid}
ga <- grid_array(nlayers=2,nrow=30,ncol=20,N=1000,mig.rate=1,admix.rate=0.1)
plot(ga)
```
In the plot, the diagonal shows arrows weighted by migration rates within a given layer,
and offdiagonals show circles with size proportional to the "admixture" migration between the relevant layers.
This looks pretty boring because all the migration rates are the same,
and isn't showing between-layer migration rates at the moment.
So we have something to look at, let's 
increase migration rates on one layer,
and then restrict each layer to a circle
centered on the two corners:
```{r zero_circles}
ga <- modify_migration(ga,layer=1,doutM=rexp(ncol(ga)*nrow(ga)),dinM=rexp(nrow(ga)*ncol(ga)))
plot(ga)
ga <- restrict_patch(ga,layer=1,xy=c(1,1),r=10)
ga <- restrict_patch(ga,layer=2,xy=c(nrow(ga),ncol(ga)),r=10)
plot(ga)
```
Note that each population also has an associated "exponential growth" rate.

We can also add barriers:
```{r grid_barrier}
ga <- grid_array(nlayers=2,nrow=30,ncol=20,N=1000,mig.rate=1,admix.rate=0.1)
ga <- add_barrier(ga, layer=1, rows=c(5,15))
ga <- add_barrier(ga, layer=2, cols=10)
plot(ga)
```

Demographies
============

Now, here's an example of stringing these together into a model,
with four model changes,
and various `dt` between them:
```{r init_dem}
dem <- demography( grid_array(nlayers=2,nrow=30,ncol=20,N=1000,mig.rate=1,admix.rate=0.1) )
dem[[1]] <- modify_migration(dem[[1]],layer=1,doutM=rexp(prod(dim(dem[[1]])[1:2])),dinM=rexp(prod(dim(dem[[1]])[1:2])))
dem <- add_to_demography( dem, dt=2, add_barrier, layer=1, rows=10 )
dem <- add_to_demography( dem, dt=1, restrict_patch, layer=1,xy=c(1,1),r=10 )
dem[[3]] <- restrict_patch(dem[[3]],layer=2,xy=c(nrow(dem[[1]]),ncol(dem[[1]])),r=10)
dem <- add_to_demography( dem, dt=2, identity )
dem[[4]] <- dem[[1]]
cat("t=",0,"\n")
plot(dem[[1]])
cat("t=",dem@t[1],"\n")
plot(dem[[2]])
cat("t=",dem@t[2],"\n")
plot(dem[[3]])
cat("t=",dem@t[3],"\n")
plot(dem[[4]])
```

We also have a function to convert these to ms command line parameters.
To check this, let's try a smaller model.
```{r ms_args}
dem <- demography( grid_array(nlayers=2,nrow=3,ncol=2,N=1000,mig.rate=1,admix.rate=0.1) )
dem[[1]] <- modify_migration(dem[[1]],layer=1,doutM=rexp(ncol(dem[[1]])),dinM=rexp(nrow(dem[[1]])))
dem <- add_to_demography( dem, dt=1, restrict_patch, layer=1,xy=c(1,1),r=1 )
dem[[2]] <- restrict_patch(dem[[2]],layer=2,xy=c(nrow(dem[[1]]),ncol(dem[[1]])),r=1)
print( msarg(dem,nsamp=cbind(row=1,col=1,layer=1,n=2)) )
```

*To-do:* make better specification of how many samples per which population.



Check on the migration rates
============================

Let's run a quick check that migration works the way we think.
Here are two scenarios: in each, there are 
two populations connected by migration, the second of which is very small (so coalescence happens instantaneously);
in the first scenario, the first population is of size 100,
in the second, it is much bigger (10,000).
We take five samples from the first, and two from the second.
Migrants only go from the first population to the second.
Since the second population is very small,
and the first is large, coalescences happen only when a migration occurs, and then immediately.
Ideally, we would plot both trees on the same scale,
but, because `ape`,
that seems to be impossible.
```{r big_small_check}
popsizes <- c(1e3,1e5)
ms.trees <- lapply( popsizes, function (popsize) {
        ga <- grid_array(
                nlayers=2,     # two layers
                nrow=1,ncol=1, # but only a 1x1 "grid" on each layer
                mig.rate=1,    # migration within each layer; does nothing here
                admix.rate=matrix(c(0,0,1,0),nrow=2),  # migration from layer 1 to layer 2, but not vice-versa
                N=c(popsize,1e-4)       # first pop is 1000x the second
            )
        ms.output <- run_ms( ga,
                nsamp=c(5,2),    # five sampled lineages from each population
                trees=TRUE,      # output the tree (in newick)
                tofile=FALSE,     # don't bother putting this in a file
                scale.migration=FALSE  # don't scale migration rates by pop sizes, to investigate what ms does by itself
            )
        cat( paste( ms.output, collapse="\n" ), "\n" )  # here's the output
        read.tree(text=ms.output[5])
    } )
layout(t(1:2))
plot(ms.trees[[1]], tip.color=ifelse(as.numeric(ms.trees[[1]]$tip.label)>5,"red","black"), main=paste("pop size:",popsizes[1]) )
add.scale.bar()
plot(ms.trees[[2]], tip.color=ifelse(as.numeric(ms.trees[[2]]$tip.label)>5,"red","black"), main=paste("pop size:",popsizes[2]) )
add.scale.bar()
```
The trees look about the same (on the same scale),
certainly not different by a factor of 100,
as you'd get if migration rates of lineages was affected by the difference in population size.

Is the same true for the source population?
Yep:
```{r source_big_small_check}
popsizes <- c(1e-2,1e-5)
ms.trees <- lapply( popsizes, function (popsize) {
        ga <- grid_array(
                nlayers=2,     # two layers
                nrow=1,ncol=1, # but only a 1x1 "grid" on each layer
                mig.rate=1,    # migration within each layer; does nothing here
                admix.rate=matrix(c(0,0,1,0),nrow=2),  # migration from layer 1 to layer 2, but not vice-versa
                N=c(1e4,popsize)       # first pop is 1000x the second
            )
        ms.output <- run_ms( ga,
                nsamp=c(5,2),    # five sampled lineages from each population
                trees=TRUE,      # output the tree (in newick)
                tofile=FALSE,     # don't bother putting this in a file
                scale.migration=FALSE  # don't scale migration rates by pop sizes, to investigate what ms does by itself
            )
        cat( paste( ms.output, collapse="\n" ), "\n" )  # here's the output
        read.tree(text=ms.output[5])
    } )
layout(t(1:2))
plot(ms.trees[[1]], tip.color=ifelse(as.numeric(ms.trees[[1]]$tip.label)>5,"red","black"), main=paste("pop size:",popsizes[1]) )
add.scale.bar()
plot(ms.trees[[2]], tip.color=ifelse(as.numeric(ms.trees[[2]]$tip.label)>5,"red","black"), main=paste("pop size:",popsizes[2]) )
add.scale.bar()
```

